---
title: "Portfolio plan: Analysis of Rush"
output:
  github_document: default
  html_document: default
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE)

# Load libraries (every time)

library(tidyverse)
library(spotifyr)
library(dplyr)
library(ggplot2)

# Set Spotify access variables (every time)

load("Corpus.RData")
load("spotify_env.RData")

set_spotify_ID()
```

__Computational Musicology 2020__

__Jesper Kuiper, 12984787__

The Canadian band [Rush](https://bit.ly/2HkT2Sx) is at its heart a rock band. However, during the extensive lifetime of Rush, its music found a lot of influence from a wide scope of genres and music styles. With a lifespan of more than 40 years and a total of 19 studio albums, a lot has changed from begin to end. 
In this portfolio, I will try to analyze the way Rush's music changed over the years, and I will try to determine different style periods accross its albums. Furthermore, I will look at how different genres apply to different eras of Rush. 

## The corpora
Of course, the most important corpus I will use is that containing the audio features of the entire discography of Rush itself. Upon initial inspection, I decided to remove the album "2112 - 40 Years Closer: A Q&A With Alex Lifeson And Terry Brown Commentary" as it contains commentary and is thus not relevant to my analysis. 
Furthermore, I looked at different genres that are said to be part of at least some era of Rush. For each of these genres I collected the audio features of a playlist that seemed most representative for that genre. I chose playlists of the following genres:

 * Rock:
    + 70s Rock Anthems
    + 80s Rock Anthems
    + 90s Rock Anthems
    + 00s Rock Anthems
 * Progressive Rock (user-made playlist)
 * Pop Rock (Pop Rock Shot)
 * Reggae (Reggae Classics)
 * New Wave (Is it New Wave?)
 * Jazz (Jazz Classics)
 * Metal (Metal Essentials)
 * Symphonic Metal (user-made playlist)
 
I chose a total of four playlists for the Rock genre, so that I can analyse whether Rush followed the general trend of rock music through the years. 

## Initial analysis
I analyzed the audio features of each studio album of Rush, to see whether there are some easily spottable trends in there. The studio albums are, in order of release date:

1. Rush (1974)
2. Fly By Night (1975)
3. Caress of Steel (1975)
4. 2112 (1976)
5. A Farewell to Kings (1977)
6. Hemispheres (1978)
7. Permanent Waves (1980)
8. Moving Pictures (1981)
9. Signals (1982)
10. Grace Under Pressure (1984)
11. Power Windows (1985)
12. Hold Your Fire (1987)
13. Presto (1989)
14. Roll The Bones (1991)
15. Counterparts (1993)
16. Test For Echo (1997)
17. Vapor Trails (2002)
19. Snakes & Arrows (2007)
20. Clockwork Angels (2012)

```{r plot, include=TRUE, eval=TRUE}
const = function(x){x[1]}
func = mean
rush_studioalbum_idx <- c(1, 2, 3, 4, 6, 7, 8, 9, 12, 13, 14, 15, 17, 19, 20, 21, 24, 27, 30)

rush_features_by_album <- rush %>% 
  filter(album_id != "3U6vR85uJOAT08DLnJhZhH") %>% 
  group_by(album_name) %>%
  summarise(
    release=const(album_release_date),
    danceability_sd=sd(danceability),
    energy_sd=sd(energy),
    speechiness_sd=sd(speechiness),
    acousticness_sd=sd(acousticness),
    liveness_sd=sd(liveness),
    valence_sd=sd(valence),
    instrumentalness_sd=sd(instrumentalness),
    instrumentalness=func(instrumentalness),
    danceability=func(danceability),
    energy=func(energy),
    speechiness=func(speechiness),
    acousticness=func(acousticness),
    liveness=func(liveness),
    valence=func(valence),
    loudness = func(loudness),
    
  ) %>% 
  arrange(release)

rush_features_by_album$album_no <- c(1:nrow(rush_features_by_album))

rush_features_by_studioalbum <- rush_features_by_album[rush_studioalbum_idx,]
rush_features_by_studioalbum$album_no <- c(1:nrow(rush_features_by_studioalbum))

plot_studioalbum_features <- ggplot(rush_features_by_studioalbum, aes(x=album_no)) +
  geom_line(aes(x=album_no, y=danceability, colour="danceability")) + 
  geom_line(aes(x=album_no, y=energy, colour="energy")) + 
  geom_line(aes(x=album_no, y=speechiness, colour="speechiness")) + 
  geom_line(aes(x=album_no, y=acousticness, colour="acousticness")) + 
  geom_line(aes(x=album_no, y=instrumentalness, colour="instrumentalness")) + 
  geom_line(aes(x=album_no, y=liveness, colour="liveness")) + 
  geom_line(aes(x=album_no, y=valence, colour="valence")) + 
  scale_x_continuous(
    name="Studio album idx",
    breaks=c(1:nrow(rush_features_by_studioalbum)), 
    labels=c(1:nrow(rush_features_by_studioalbum)),
    limits=c(1,nrow(rush_features_by_studioalbum))) +
  scale_y_continuous(
    name="Value",
    breaks=c(0:10)/10, 
    labels=c(0:10)/10,
    limits=c(0,1)) +
  ggtitle("Mean of music features per studio album") +
  labs(color="Property")
plot_studioalbum_features
```

This initial plot looks promising; we can see at least the _energy_, _valence_ and _danceability_ following some kind of a pattern here. Interestingly, we can see a dip in _energy_ and a peak in both _danceability_ and _valence_ at the 13th studio album. It seems like Rush tried to do something different in that album. This could indeed be what happened, as this album - Presto - wasn't received very well by the general public. It seems Rush learned from this, as, according to my plot, they seem to return to their 'roots' in their later albums. 

```{r tempoplot}
rush_tempoloudns_by_album <- rush %>% 
  filter(album_id != "3U6vR85uJOAT08DLnJhZhH") %>% 
  group_by(album_name) %>%
  summarise(
    release=const(album_release_date),
    tempo=mean(tempo),
    loudness=mean(loudness)
  ) %>% 
  arrange(release)

rush_tempo_by_studioalbum <- rush_tempoloudns_by_album[rush_studioalbum_idx,c("album_name", "release", "tempo")]
rush_loudn_by_studioalbum <- rush_tempoloudns_by_album[rush_studioalbum_idx,c("album_name", "release", "loudness")]

plot_tempo_rush <- ggplot(rush_tempo_by_studioalbum, aes(x=c(1:nrow(rush_tempo_by_studioalbum)), y=tempo)) + 
    geom_line()
plot_loudness_rush <- ggplot(rush_loudn_by_studioalbum, aes(x=c(1:nrow(rush_loudn_by_studioalbum)), y=loudness)) + 
    geom_line()
plot_loudness_rush
    
```

[^1]: The link may not work; this is because for some reason Github inserts a '/' before a bracket. 